Stage-Splitting a Modal Language

### 2014-09-30 paper/example.tex
fix unicode in pdf bookmarks
made changes from meeting with Umut

\section {Example: Staged Quickselect in \lang}
\label{sec:staging}

@@

As an example of an algorithm amenable to staging techniques,
consider the implementation of quickselect shown in Figure~2a.
Quickselect is an $O(n)$ (expected time) algorithm for finding the $k$-th largest element in a randomly ordered length-$n$ list.
It does so by partitioning the input list into sublists containing elements less than and greater than the head element, 
then recurring on the sublist containing the desired element.

Importantly, one property of quickselect is that the partitioning of the list, even through its recursive calls, depends only on the values of list elements, not on the value of $k$ 
\footnote{There is a control-dependence on $k$ -- that is $k$ determines what splits are made but not how they are made -- we ignore this for now.}.  Therefore, if quickselect was to be evaluated many times on the same list, this independence admits a familiar optimization where the algorithm is split into two functions: one which precomputes \emph{all partitions} of the input list needed for any value of $k$ (resulting in a binary search tree---see \texttt{qSelect1} in Figure~2c),
and another which, given $k$, uses the tree to perform accelerated lookup of the $k$-th element (essentially performing binary search of a sorted list---see \texttt{qSelect2}).
%One advantage of splitting the algorithm into two phases is that the search tree can be constructed for a list once
%and then used to accelerate many future lookups for any value of $k$.

>>>>

\section {Example: Staged Quickselect in \texorpdfstring{\lang}{λ12}}
\label{sec:staging}

@@

As an example of an algorithm amenable to staging techniques,
consider the implementation of quickselect shown in Figure~2a.
Quickselect is an algorithm for finding the $k$-th largest element in a list.
It works by partitioning the input list into sublists containing elements less than and greater than the head element, 
then recurring on the sublist containing the desired element.

One notable property of quickselect is that the partitioning of the list, even through its recursive calls, depends only on the values of list elements, not on the value of $k$ 
\footnote{There is a control-dependence on $k$ -- that is $k$ determines what splits are made but not how they are made -- we ignore this for now.}.  Therefore, if quickselect was to be evaluated many times on the same list, this independence admits a familiar optimization where the algorithm is split into two functions: one which precomputes \emph{all partitions} of the input list needed for any value of $k$ (resulting in a binary search tree---see \texttt{qSelect1} in Figure~2c),
and another which, given $k$, uses the tree to perform accelerated lookup of the $k$-th element (essentially performing binary search of a sorted list---see \texttt{qSelect2}).

The benefits are substantial: for a randomly ordered list of length $n$, 
the original \texttt{qSelect} takes expected $O(n)$ time per lookup, 
whereas \texttt{qSelect2} takes $O(\log n)$ time.
Since \texttt{qSelect1} does not depend on $k$, its $O(n\log n)$ cost can be amortized over many lookups.
%One advantage of splitting the algorithm into two phases is that the search tree can be constructed for a list once
%and then used to accelerate many future lookups for any value of $k$.

### 2014-09-30 paper/relatedwork.tex
made changes from meeting with Umut

@@

@@

[Discuss Yong's recent paper here.  It does some pretty sophisticated binding time analysis, with a somewhat straightforward splitting after that.  They have the same many-to-one use case as Spark, but syntactic restrictions prevent sum types on the boundary, sort of.  If we wanted to faithfully represent their system in ours, we would need some mechanisms for abstraction over stage, which we do not have.]

Davies (\cite{davies96}) explored the connection between linear temporal logic and its corresponding type system [circle](which we adapted into [circle sub 2]), and showed the equivalence between [circle] and existing systems for binding time analysis. That work provided $\beta$ and $\eta$ rules for the next and prev operators, but did not consider a full dynamic semantics for the whole language. Whereas [name of our type system] is appropriate for stage-splitting and partial evaluation, \cite{davies01} provides a similar system, [square], that is appropriate for meta-programming.  The main difference is that terms inside a [prev] operator do not see any stage-2 bindings declared outside of it.  They note that where [circle] corresponds to the non-branching temporal logic, [square] corresponds to the branching version.

>>>>

\subsection{Davies}

\cite{davies96} explored the connection between linear temporal logic and its corresponding type system (which is nearly identical to that for \lang), 
and showed the equivalence between that type system and existing one for binding time analysis. 
That work also provided a metaprogramming-style duplicating dynamic semantics, which we purposely avoided.
That said, the dynamics of \cite{davies96} is otherwise very similar to ours.
In fact, if we change the $\next$ and $\prev$ rules of ours to 
\begin{mathpar}
\infer {\diaone {\next~e}{\cdot,\next~q}} {\diatwo e q} \and
\infer {\diatwo{\prev~e}{v}} {\diaone e {\cdot,\next~v}} 
\end{mathpar}
and treat $\next~q$ as a partial value for any residual $q$,
then our semantics becomes rule-for-rule isomorphic to that from \cite{davies96}. This essentially bypasses the
environment bookkeeping in $\redonesym$, by inlining residuals instead of
hoisting them in \verb|let|-bindings.
From this it's clear that the semantics of \cite{davies96} and ours always produce the same value when they both terminate.
{\em However because \lang's semantics dictate that a reified residual will always be evaluated, regardless of whether its result is consumed, \lang\ programs terminate strictly less often than that of \cite{davies96}}.

\subsection{Specialization}

@@

\subsection{Graphics Systems}

@@

### 2014-09-30 paper/semantics.tex
made changes from meeting with Umut

Intuitively, the solution is to designate \verb|next {f 20}| as a value
\emph{in stage \bbone}, even though it requires additional evaluation in stage
\bbtwo. Therefore, the pair evaluates to
\begin{lstlisting}
#2 (next {f 20}, 5)
\end{lstlisting}
and the projection, in turn, immediately evaluates to \verb|5|.

Now consider a more complex example where stage~\bbone\ evaluation must
substitute such a incompletely-evaluated expression. The following
stage-\bbtwo\ term has type $\rmint$:
\begin{lstlisting} 
prev{
  let x = (next {f 20}, 3+4) in
  next{ prev{#1 x} * prev{#1 x} * hold{#2 x} }
}
\end{lstlisting}
Again, this term does not fully reduce at stage~\bbone, because it depends on
the value of \verb|f 20|, which is not reduced until stage \bbtwo.

If we simply treat \verb|(next {f 20}, 7)| as a value during stage~\bbone, and
substitute it for the three occurrences of \verb|x| in the body of the
\verb|let| expression, the stage \bbone\ result would be
\begin{lstlisting} 
prev{
  next{ 
    prev{next {f 20}} * prev{next {f 20}} * 7 
  }
}
\end{lstlisting}
Finally, the $\prev$s and $\next$s cancel to yield the residual:
\begin{lstlisting} 
(f 20) * (f 20) * 7
\end{lstlisting}
But now stage \bbtwo\ must compute the expensive function call \verb|f 20| twice! 
There are some systems, typically called {\em metaprogramming}, 
where this duplication would be considered intended behavior.
For our applications, we find the duplication undesirable and avoid it.
Accordingly, instead of treating \verb|next {f 20}| as a
value and substituting it into the body of the \verb|let|, we instead bind
\verb|f 20| to a stage~\bbtwo\ variable $\mathtt{\hat y}$, and substitute that
variable.

More precisely, when evaluating the \verb|next|, we construct an explicit
substitution $\mathtt{[\hat y\mapsto f~20]}$ binding its old contents to the
fresh variable $\mathtt{\hat y}$. This substitution is placed at the top of the
containing $\prev$ block:
\begin{lstlisting} 
prev {
[yhat|->f 20]
  let x = (next{yhat}, 7) in
  next{prev{#1 x} * prev{#1 x} *  hold{#2 x}}
}
\end{lstlisting}
%As a convention, we render the new variable with a %stylish and fashionable
%hat.  
We proceed with stage~\bbone\ evaluation, which now duplicates the variable
$\mathtt{\hat y}$ rather than the expression \verb|f 20|.
\begin{lstlisting} 
prev {
[yhat|->f 20]
  next{
    prev{#1 (next {yhat}, 7)} * 
    prev{#1 (next {yhat}, 7)} *
    hold{#2 (next {yhat}, 7)}
  }
}
\end{lstlisting}
To evaluate the large $\next$, we must first \emph{speculate} inside of it, to
evaluate all the stage~\bbone\ terms contained within $\prev$s. These
necessarily reduce to $\next$ expressions which cancel the $\prev$s, resulting
in:
\begin{lstlisting} 
prev {
[yhat|->f 20]
    next{ yhat * yhat * 7 }
}
\end{lstlisting}
Once again, we lift the contents of the $\next$ into a substitution:
\begin{lstlisting} 
prev {
[yhat|->f 20]
[zhat|->yhat*yhat*7]
    next{ zhat }
}
\end{lstlisting}
Finally, when evaluating the outer $\prev$, we must {\em reify} the contained substitutions into let statements, yielding
\begin{lstlisting} 
let yhat = f 20 in
let zhat = yhat * yhat * 7 in z
\end{lstlisting}

@@

The $\redonesym$ judgment takes an open stage \bbone\ term to a {\em future
environment} $\xi$ and a partial value $v$.  The future environment implements
the explicit substitutions in our example---it maps fresh stage \bbtwo\
variables (which may appear inside $\next$ blocks in $v$) to
residuals. For non-staging features of \lang\ (\ref{fig:diaSemanticsCore}),
$\redonesym$ is essentially standard call-by-value evaluation, and gathers the
subterms' future environments into a single one.

@@

Within speculation lies a subtle---if perhaps unintuitive---feature.  
Observe that speculation will traverse into both branches of a stage-\bbtwo\ {\tt if} or {\tt case} 
statement in its search for stage~\bbone\ code. 
Thus the evaluation of that stage~\bbone\ code will occur {\em regardless of the eventual value of the predicate},
and so a term like 
\begin{lstlisting} 
next{
  if true 
  then hold{1+2} 
  else prev{spin() (* loops forever *)}
}
\end{lstlisting}
will fail to evaluate at stage \bbone.
This behavior is why the judgment is named ``speculation."
Although it may seem undesirable, this speculative behavior is a deliberate feature.
Indeed it is critical to the quickselect example, 
wherein all three branches of the $n$-$k$ comparison are evaluated at stage~\bbone.

@@

As an optimization, we can include the special-case rule,
\begin{mathpar}
\inferdiaone [hat] {\red {\next~\hat y}{\cdot,\next~\hat y}}{\cdot}
\end{mathpar}
to avoid one-for-one variable bindings in the residual.
We used this implicitly in the example in the previous section.

\TODO [below here will get moved to related work]

If we change the $\next$ and $\prev$ rules to 
\begin{mathpar}
\infer {\diaone {\next~e}{\cdot,\next~q}} {\diatwo e q} \and
\infer {\diatwo{\prev~e}{v}} {\diaone e {\cdot,\next~v}} 
\end{mathpar}
and treat $\next~q$ as a partial value for any residual $q$,
then our semantics becomes rule-for-rule isomorphic to that from \cite{davies96}. This essentially bypasses the
environment bookkeeping in $\redonesym$, by inlining residuals instead of
hoisting them in \verb|let|-bindings.
From this it's clear that the semantics of \cite{davies96} and ours always produce the same value when they both terminate.
{\em Because \lang's semantics dictate that a reified residual will always be evaluated, regardless of whether its result is consumed, \lang\ programs terminate strictly less often than that of \cite{davies96}}.

>>>>

Intuitively, the solution is to designate \verb|next {f 20}| as a value
\emph{in stage \bbone}, even though it requires additional evaluation in stage
\bbtwo. Therefore, the pair evaluates to
\begin{lstlisting}
(next {f 20}, 5)
\end{lstlisting}
and the projection, in turn, immediately evaluates to \verb|5|.

Now consider a more complex example where stage~\bbone\ evaluation must
substitute such a incompletely-evaluated expression. The following
stage-\bbtwo\ term has type $\rmint$:
\begin{lstlisting} 
prev{
  let x = (next {f 20}, 3+4) in
  next{ prev{#1 x} * prev{#1 x} * hold{#2 x} }
}
\end{lstlisting}
As with the simpler example, this term does not fully reduce at stage~\bbone, 
because it depends on the value of \verb|f 20|, which is not reduced until stage \bbtwo.

Again treating \verb|(next {f 20}, 7)| as a value during stage~\bbone, we
substitute it for the three occurrences of \verb|x| in the body of the
\verb|let| expression, yielding
\begin{lstlisting} 
prev{
  next{ prev{#1 (next {f 20}, 7)} 
      * prev{#1 (next {f 20}, 7)} 
      * hold{#2 (next {f 20}, 7)} }
}
\end{lstlisting}
To then evaluate the large $\next$, we must first search it for stage~\bbone\ subterms to reduce in place.
This search process is called {\em speculation}.
In our example this means reducing all three projections to give,
\begin{lstlisting} 
prev{
  next{ 
    prev{next {f 20}} * prev{next {f 20}} * 7 
  }
}
\end{lstlisting}
Since all stage~\bbone\ subterms are reduced, the $\prev$s and $\next$s cancel to yield the final residual:
\begin{lstlisting} 
(f 20) * (f 20) * 7
\end{lstlisting}
But now stage \bbtwo\ must compute the expensive function call \verb|f 20| twice! 
There are some systems, typically called {\em metaprogramming}, 
where this duplication would be considered intended behavior.
For our applications, we find the duplication undesirable and avoid it.

To get this behavior, we do not treat \verb|next {f 20}| as a value to be substituted, 
but instead bind \verb|f 20| to a fresh stage~\bbtwo\ variable $\mathtt{\hat y}$
and then substitute that variable.

More precisely, when evaluating the \verb|next|, we construct an explicit
substitution $\mathtt{[\hat y\mapsto f~20]}$ binding its old contents to the
fresh variable $\mathtt{\hat y}$. This substitution is placed at the top of the
containing $\prev$ block:
\begin{lstlisting} 
prev {
[yhat|->f 20]
  let x = (next{yhat}, 7) in
  next{prev{#1 x} * prev{#1 x} *  hold{#2 x}}
}
\end{lstlisting}
%As a convention, we render the new variable with a %stylish and fashionable
%hat.  
We proceed with stage~\bbone\ evaluation, which now duplicates the variable
$\mathtt{\hat y}$ rather than the expression \verb|f 20|.
\begin{lstlisting} 
prev {
[yhat|->f 20]
  next{
    prev{#1 (next {yhat}, 7)} * 
    prev{#1 (next {yhat}, 7)} *
    hold{#2 (next {yhat}, 7)}
  }
}
\end{lstlisting}
We then speculate and reduce the projections to get:
\begin{lstlisting} 
prev {
[yhat|->f 20]
    next{ yhat * yhat * 7 }
}
\end{lstlisting}
Once again, we lift the contents of the $\next$ into a substitution:
\begin{lstlisting} 
prev {
[yhat|->f 20]
[zhat|->yhat*yhat*7]
    next{ zhat }
}
\end{lstlisting}
Finally, when evaluating the outer $\prev$, we must {\em reify} the contained substitutions into let statements, yielding
\begin{lstlisting} 
let yhat = f 20 in
let zhat = yhat * yhat * 7 in z
\end{lstlisting}

@@

The $\redonesym$ judgment takes an stage \bbone\ term\footnote{
The input to $\redonesym$ may be open on stage \bbtwo\ variables,
but the type system ensures that those must occur under a $\next$.
Consequently $\redonesym$ will never directly encounter a variable.
} to a {\em future environment} $\xi$ and a partial value $v$.  
The future environment implements
the explicit substitutions in our example---it maps fresh stage \bbtwo\
variables (which may appear inside $\next$ blocks in $v$) to
residuals. For non-staging features of \lang\ (\ref{fig:diaSemanticsCore}),
$\redonesym$ is essentially standard call-by-value evaluation, and gathers the
subterms' future environments into a single one.

@@

Within speculation lies a subtle---if perhaps unintuitive---feature.  
Observe that speculation will traverse into both branches of a stage-\bbtwo\ {\tt if} or {\tt case} 
statement in its search for stage~\bbone\ code. 
Thus the evaluation of that stage~\bbone\ code will occur {\em regardless of the eventual value of the predicate},
and so a term like 
\begin{lstlisting} 
next{
  if true 
  then hold{1+2} 
  else prev{spin() (* loops forever *)}
}
\end{lstlisting}
will fail to evaluate at stage \bbone.
This behavior is why the judgment is named ``speculation."
Although it may seem undesirable, this speculative behavior is a deliberate feature.
Indeed it is critical to the quickselect example, 
wherein all three branches of the $n$-$k$ comparison are evaluated at stage~\bbone.
As always, it is the programmer's responsibility to ensure that her staging annotations 
do not cause a program to loop forever.

@@

As an optimization, we can include the special-case rule,
\begin{mathpar}
\inferdiaone [hat] {\red {\next~\hat y}{\cdot,\next~\hat y}}{\cdot}
\end{mathpar}
to avoid one-for-one variable bindings in the residual.
